---
title: "206 Project Appendix B"
author: "Longbei Xu, Gu Gong, Yichu Chen"
date: "12/2/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Appendix B

```{r, echo=FALSE}
library(reshape2)
library(ggplot2)
library(GGally)
library(RColorBrewer)
library(MASS)
library(leaps)
library(car)

set.seed(123)
```


```{r, results='hide'}
# import data
plasma <- read.delim("Plasma.txt", stringsAsFactors=TRUE)

# check missing value
colSums(is.na.data.frame(plasma))

# check data type
lapply(plasma, class)
```

```{r}
# numerical and categorical sub.dataframe
num_data <- plasma[,unlist(lapply(plasma, is.numeric))]
cat_data <- plasma[,!unlist(lapply(plasma, is.numeric))]
cat.response <- cbind(plasma[,c('BETAPLASMA','RETPLASMA')], cat_data)

# column names
num_names <- names(num_data)
cat_names <- names(cat_data)
```

```{r, fig.width=12, fig.height=10, message=FALSE, fig.show='hide'}
lowerF <- function(data, mapping, method = "lm") {
  p <- ggplot(data = data, mapping = mapping) +
    # scatter plot
    geom_point(alpha=0.3) +
    # smoothed regression line
    geom_smooth(method = method, color = "red")
  p
}
ggpairs(num_data, lower = list(continuous = wrap(lowerF, method= "lm")))
```

```{r, fig.show='hide'}
# correlation heatmap
ggcorr(num_data, method = c("everything", "pearson"), cex=2)+labs(title="Correlation Heatmap for Numerical Variables")
```

```{r, fig.show='hide'}
# histogram and boxplots for continuous variables
num_long <- melt(num_data)

ggplot(num_long,aes(x = value)) + 
    facet_wrap(~variable,scales = "free") + 
    geom_histogram(aes(y = ..density..), color = 2, fill = "white") +
    geom_density(lwd = 0.6,
               linetype = 2,
               color = 1,
               fill=2,
               alpha=0.1)

ggplot(num_long)+
  facet_wrap(~variable, scales='free')+
  geom_boxplot(aes(y= value)) 
```

```{r, warning=FALSE, fig.width=16, fig.show='hide'}
# pie charts for categorical variables

par(mfrow=c(1,3))
tab_vit <- table(cat_data$VITUSE)
pct_vit <- round(100*tab_vit/nrow(cat_data))
label_vit <- paste(levels(cat_data$VITUSE), pct_vit)
label_vit <- paste(label_vit, "%", sep=" ")
pie(tab_vit, border="white", labels=label_vit, col= brewer.pal(length(levels(cat_data$VITUSE)), "Pastel1"), main="Vitamin Use")

tab_sex <- table(cat_data$SEX)
pct_sex <- round(100*tab_sex/nrow(cat_data))
label_sex <- paste(levels(cat_data$SEX), pct_sex)
label_sex <- paste(label_sex, "%", sep=" ")
pie(tab_sex, border="white", labels=label_sex, col= brewer.pal(length(levels(cat_data$SEX)), "Pastel1"), main="Sex")

tab_smoke <- table(cat_data$SMOKSTAT)
pct_smoke <- round(100*tab_smoke/nrow(cat_data))
label_smoke <- paste(levels(cat_data$SMOKSTAT), pct_smoke)
label_smoke <- paste(label_smoke, "%", sep=" ")
pie(tab_smoke, border="white", labels=label_smoke, col= brewer.pal(length(levels(cat_data$SMOKSTAT)), "Pastel1"), main="Smoking Status")


```
```{r, fig.width=8, fig.show='hide'}
# boxplot for categorical variables vs. response variables

library(patchwork)

par(mfrow=c(1,2))

p1 <- ggplot(cat.response, aes(x=VITUSE)) +
      geom_boxplot(position = "dodge2", aes(y=BETAPLASMA)) +
      labs(title="Beta-Plasma VS. Vitamin Use")

  
p2 <- ggplot(cat.response, aes(x=VITUSE)) +
      geom_boxplot(position = "dodge2", aes(y=RETPLASMA)) +
      labs(title="Retinol-Plasma VS. Vitamin Use")

p3 <- ggplot(cat.response, aes(x=SEX)) +
      geom_boxplot(position = "dodge2", aes(y=BETAPLASMA)) +
      labs(title="Beta-Plasma VS. Sex")

p4 <- ggplot(cat.response, aes(x=SEX)) +
      geom_boxplot(position = "dodge2", aes(y=RETPLASMA)) +
      labs(title="Retinol-Plasma VS. Sex")

p5 <- ggplot(cat.response, aes(x=SMOKSTAT)) +
      geom_boxplot(position = "dodge2", aes(y=BETAPLASMA)) +
      labs(title="Beta-Plasma VS. Smoking Status")

p6 <- ggplot(cat.response, aes(x=SMOKSTAT)) +
      geom_boxplot(position = "dodge2", aes(y=RETPLASMA)) +
      labs(title="Retinol-Plasma VS. Smoking Status")


p1+p3+p5+p2+p4+p6+plot_layout(ncol = 3)

   
```

```{r}
set.seed(123)

# train-test split
N <- nrow(plasma)
pool <- c(1:N)

train_idx <- sample(pool, size=ceiling(N*0.9))
valid_idx <- pool[-train_idx]

# split for 2 responses
betaplasma <- plasma[, !names(plasma)%in%c("RETPLASMA")]
retplasma <- plasma[, !names(plasma)%in%c("BETAPLASMA")]

beta_train <- betaplasma[train_idx,]
beta_valid <- betaplasma[valid_idx,]

ret_train <- retplasma[train_idx,]
ret_valid <- retplasma[valid_idx,]
```


## Beta-carotene plasma concentration

### Initial Checking
```{r}
#First-order full model before transformation (used for Box-Cox and assumption checking)
beta.reg <- lm(BETAPLASMA +1~ ., data=beta_train)
# summary(beta.reg)
```

```{r, fig.show='hide'}
# assumption checking for beta-carotene
par(mfrow=c(1,2))
plot(beta.reg, which=1)
plot(beta.reg, which=2)
```

```{r, fig.show='hide'}
# Box-Cox procedure for beta-carotene
library(MASS)
bc = boxcox(beta.reg)
```

```{r}
# First-order full model with transformed Y
beta.reg <- lm(log(BETAPLASMA+1) ~ ., data=beta_train)
```

```{r, fig.show='hide'}
# initial assumption checking for beta carotene
par(mfrow=c(1,2))
plot(beta.reg, which=1)
plot(beta.reg, which=2)
```

### Candidate Model (without interaction) for beta-carotene

```{r, results='hide'}
#first-order model stepwise selection using AIC
step.reg.beta <- stepAIC(beta.reg, direction="both", trace=0)
summary(step.reg.beta)

# model AIC
AIC(step.reg.beta)
```

```{r, fig.show='hide'}
#residuals for first-order model (based on stepwise selection)
{plot(step.reg.beta, which=1)
plot(step.reg.beta, which=2)}
```

### Final Model (with interaction) for beta-carotene

```{r, results='hide'}
# interaction model for beta-carotene (stepwise AIC)
step.reg <- stepAIC(lm(log(BETAPLASMA+1)~1, data=beta_train), scope = ~(AGE+SEX+SMOKSTAT+QUETELET+VITUSE+CALORIES+FAT+FIBER+ALCOHOL+CHOLESTEROL+BETADIET+RETDIET)^2, direction="both", trace=0)

summary(step.reg)

# model AIC
AIC(step.reg)

# anova table
step.reg$anova
```


## Retinol plasma concentration

### Initial Checking
```{r, fig.show='hide'}
#First-order before transformation 
ret.reg <- lm(RETPLASMA ~ ., data=ret_train)
# summary(ret.reg)

# before transformation
par(mfrow=c(1,2))
plot(ret.reg, which=1)
plot(ret.reg, which=2)
```

```{r, fig.show='hide'}
# Box-Cox transformation
library(MASS)
bc1 =boxcox(ret.reg, data=ret_train)
```
```{r}
#First-order with transformation
ret.reg <- lm(log(RETPLASMA) ~ ., data=ret_train)
# summary(ret.reg)
```

```{r, fig.show='hide'}
# after transformation: assumption checking
par(mfrow=c(1,2))
plot(ret.reg, which=1)
plot(ret.reg, which=2)
```

### Candidate Model (without interaction) for retinol
```{r, results='hide'}
#first-order stepwise using AIC

ret.reg0 <- lm(log(RETPLASMA) ~ 1, data=ret_train)
ret.reg <- stepAIC(ret.reg0, 
                   scope = ~(AGE+SEX+SMOKSTAT+QUETELET+VITUSE+CALORIES+FAT+FIBER+ALCOHOL+CHOLESTEROL+BETADIET+RETDIET), direction="both", trace=0)
summary(ret.reg)

AIC(ret.reg)
```

```{r, fig.show='hide'}
#residuals for assumption checking
{plot(ret.reg, which=1)
plot(ret.reg, which=2)}
```


### Final Model (with interaction) for retinol
```{r, results='hide'}

none_mod<- lm(RETPLASMA~1, data=ret_train) ##model with only intercept

full_mod<-lm(RETPLASMA~(.)^2,data=ret_train)
# summary(full_mod)

library(MASS)
#forward selection based on AIC: 
stepRET.aic<-stepAIC(none_mod, scope=list(upper=full_mod, lower = ~1), direction="both", k=2, trace = FALSE)
summary(stepRET.aic)

stepRET.aic$anova

AIC(stepRET.aic)
```




## Model validation

### Beta-Carotene

```{r}
step.reg.beta <- stepAIC(lm(log(BETAPLASMA+1)~1, data=beta_train), scope = ~(AGE+SEX+SMOKSTAT+QUETELET+VITUSE+CALORIES+FAT+FIBER+ALCOHOL+CHOLESTEROL+BETADIET+RETDIET)^2, direction="both", trace=0)

# train set SSE
sse_t <- sum(step.reg.beta$residuals^2)
```


#### Variance Inflation Factor
```{r, results='hide'}
# VIF
vif(step.reg.beta)
```

#### MSPE
```{r, results='hide'}
# MSPE
y.hat.beta <- predict(step.reg.beta, beta_valid)
MSPE.beta <- mean((log(beta_valid$BETAPLASMA+1) - y.hat.beta)^2)
MSPE.beta
```
```{r, results='hide'}
# SSE/n
n=N*0.7
sse.beta_train <- sum(step.reg.beta$residuals^2)
sse.beta_train/n
```

#### Outlying Y
```{r, fig.show='hide'}
# final model for beta-carotene, fit1
fit1<- lm(log(BETAPLASMA+1)~ QUETELET + VITUSE + CHOLESTEROL + BETADIET + FIBER + AGE + VITUSE:BETADIET, data=beta_train)
e1<-fit1$residuals  ##ordinary residuals 
h1<-influence(fit1)$hat ##diagonals of the hat matrix: a.k.a. leverage values 
de1<-e1/(1-h1) ##deleted residuals 
{plot(e1,de1, xlab="residuals", ylab="deleted residuals")
abline(0,1)}
```

```{r, results='hide'}
summary(h1)
```

```{r, results='hide'}
library(MASS)
stu.res.del1 <- studres(fit1)
head(sort(abs(stu.res.del1), decreasing=TRUE))
```

```{r}
# Bonferroni-corrected threshold for outlying Y
qt(1-.1/(2*n), n-10-1)
```

```{r, results='hide'}
h1 <- influence(fit1)$hat
p1 <- 10
sort(h1[which(h1>2*p1/n)], decreasing = TRUE)
```


#### Cook's Distance
```{r}
# Cook's distance
res1 <- fit1$residuals
mse1 <- anova(fit1)["Residuals", 3]
cook.d1 <- res1^2*h1/(p1*mse1*(1-h1)^2)
```


#### Influencial Points based on Cook's Distance
```{r, results='hide'}
# ranking of each observation based on Cook's Distance
sort(cook.d1[which(cook.d1>4/(n-p1))], decreasing = TRUE)
```
```{r, fig.show='hide'}
plot(fit1, which=4)
```
```{r, fig.show='hide'}
plot(fit1, which=5)
```

#### Evaluation on effect of influencial point
```{r, results='hide'}
fit.257<-lm(log(BETAPLASMA+1)~  QUETELET + VITUSE + CHOLESTEROL + BETADIET + FIBER + AGE + VITUSE:BETADIET, data=beta_train, subset=setdiff(rownames(beta_train), "257")) ##exclude case 257
rbind(fit1$coefficients,fit.257$coefficients) ##compare fitted regression coefficients
```

```{r, fig.show='hide'}
{plot(fit1$fitted.value, predict(fit.257, beta_train[,c("QUETELET","CHOLESTEROL","BETADIET","FIBER","VITUSE","AGE")]), xlab="fitted values using all cases", ylab="fitted values without using case 257") ## compare fitted values
abline(0,1)}
```



### Retinol

```{r, results='hide'}
# the final model
stepRET.aic
# SSE
sse.ret_train <- sum(stepRET.aic$residuals^2)
```

#### Variance Inflation Factor
```{r, results='hide'}
# VIF
vif(ret.reg)
```

#### MSPE
```{r, results='hide'}
# MSPE
newdata2 <- ret_valid
y.hat.ret <- predict(stepRET.aic, newdata2)
MSPE2 <- mean((log(ret_valid$RETPLASMA+1) - y.hat.ret)^2)
MSPE2
```
```{r, results='hide'}
# SSE/n
n=N*0.7
sse.ret_train/n
```

#### Outlying Y

```{r, fig.show='hide'}
# final model for retinol, fit2
fit2<- lm(log(RETPLASMA+1)~ ALCOHOL + AGE +  CHOLESTEROL + SEX + CHOLESTEROL:SEX, data = ret_train)
e2<-fit2$residuals ##ordinary residuals 
h2<-influence(fit2)$hat ##diagonals of the hat matrix: a.k.a. leverage values 
de2<-e2/(1-h2) ##deleted residuals 
{plot(e2,de2, xlab="residuals", ylab="deleted residuals")
abline(0,1)}
```

```{r, results='hide'}
summary(h2)
```

```{r, results='hide'}
library(MASS)
stu.res.del2 <- studres(fit2)
head(sort(abs(stu.res.del2), decreasing=TRUE))
```

```{r, results='hide'}
# Bonferroni-corrected threshold
qt(1-.1/(2*n), n-6-1)
```

```{r, results='hide'}
p2 <- 6
sort(h2[which(h2>2*p2/n)], decreasing = TRUE)
```

#### Cook's Distance
```{r}
# Cook's Distance
res2 <- fit2$residuals
mse2 <- anova(fit2)["Residuals", 3]
cook.d2 <- res2^2*h2/(p2*mse2*(1-h2)^2)

```


#### Influencial Points based on Cook's Distance
```{r, results='hide'}
# ranking of each observation based on Cook's Distance
sort(cook.d2[which(cook.d2>4/(n-p2))], decreasing = TRUE)
```

```{r, fig.show='hide'}
plot(fit2, which=4)
```
```{r, fig.show='hide'}
plot(fit2, which=5)
```

#### Evaluation on effect of influencial point
```{r, results='hide'}
fit.27649<-lm(log(RETPLASMA+1)~ ALCOHOL + AGE +  CHOLESTEROL + SEX + CHOLESTEROL:SEX, data = ret_train, subset=setdiff(rownames(ret_train), "276,49")) ##exclude case 276,49
rbind(fit2$coefficients,fit.27649$coefficients) ##compare fitted regression coefficients
```


```{r, fig.show='hide'}
{plot(fit2$fitted.value, predict(fit.27649, ret_train[,c("ALCOHOL" , "AGE" , "SEX" , "CHOLESTEROL")]), xlab="fitted values using all cases", ylab="fitted values with0out using case 308") ## compare fitted values
abline(0,1)}
```


### Final Evaluation on Goodness-of-fit
```{r, fig.show='hide'}
# goodness of fit for final models
par(mfrow=c(1,2))
plot(fit1, which=1)
plot(fit1, which=2)

par(mfrow=c(1,2))
plot(fit2, which=1)
plot(fit2, which=2)
```

### Model Summary
```{r}
# beta-carotene
summary(fit1)
```

```{r}
# retinol
summary(fit2)
```

### ANOVA table
```{r}
# beta-carotene
anova(fit1)
```

```{r}
# retinol
anova(fit2)
```


```{r}
vif1.test = lm(log(BETAPLASMA+1)~ QUETELET + VITUSE + CHOLESTEROL + BETADIET + FIBER + AGE, data=beta_train)
vif(vif1.test)
```

```{r}
vif2.test = lm(log(RETPLASMA+1)~ ALCOHOL + AGE +  CHOLESTEROL + SEX, data=ret_train)
vif(vif2.test)
```

